---
title: "Advanced Simulation Summative Coursework"
subtitle: 'MSc in Statistics 2025/26, Imperial College London'
author: "06051710"
format:
  html:
    toc: true
    highlight: tango
    self-contained: true
    df-print: paged
  pdf: default
format-links: false
editor: 
  markdown: 
    wrap: 72
---

```{=html}
<style type="text/css">
h1{
  font-size: 24pt;
}
h2{
  font-size: 18pt;
}
body{
  font-size: 12pt;
}
</style>
```

```{r setup, include = FALSE, tidy=TRUE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
include_solutions <- TRUE
```

```{r setup2, include=FALSE, tidy=TRUE}
set.seed(8)
require(rmarkdown)
require(knitr)
require(kableExtra)
# Put any library imports and other preamble here.
```

# Filtering and smoothing for a nonlinear state space model

```{r}
#| echo: False
#Simulating the dataset
x <- rep(0,201)
y <- rep(0,201)
x[1] <- rnorm(1)
y[1] <- (x[1]^2)/20 + rnorm(1)

for (i in 2:201){
  x[i] <- 0.5*x[i-1] + (25*x[i-1])/(1+(x[i-1])^2) + 8*cos(1.2*(i-1))+rnorm(1, mean=0, sd=sqrt(0.1))
  y[i] <- (x[i]^2)/20 + rnorm(1)
}
```

## The SIR particle filter

### 1. Sampling

For this problem, I chose as my importance density $q(x_n|y_n, x_{n-1})$ an approximation of the optimal proposal (that is $p(x_n|y_n, x_{n-1})$), by linearisations similar to the extended Kalman Filter (EKF). Although a bootstrap proposal would be easier, I felt the density $g(y_n|x_n)$ was relatively informative (or 'peaky'), which meant a more informative proposal seemed appropriate. Additionally, the notes suggest the EKF can give reasonable answers for models where the non-linearity is independent of the noise, which is the case for our model, as neither $V_n$ or $W_n$ appear in non-linear functions. 

$X_n$ can be linearly approximated as  $X_n = a_n + A_nX_{n-1} + B_nZ_n$ (where $Z_n \sim N(0,1)$) and $Y_n$ can be approximated as $Y_n = c_n +C_nX_n + W_n$ (where $W_n \sim N(0,1)$), both by a classic Taylor expansion. This leads to $B_n = \sqrt{0.1}$, $C_n = \frac{\psi_{\theta}(X_{n-1},0)}{10}$ and $D_n=1$  (all other variables are not relevant for the proposal). The approximation of the optimal proposal^[this result is taken from the notes] then takes the form $q(x_n|y_n, x_{n-1}) \propto N(m_n(x_{n-1}, y_n), S_n(x_{n-1}))$, where we have:

$$
\begin{aligned}
S_n^{-1}(x_{n-1}) &= (B_n B_n^T)^{-1} + C_n^T (D_n D_n^T)^{-1} C_n \\
S_n^{-1}(x_{n-1}) &= 10 + \frac{\psi_{\theta}(x_{n-1},0)^2}{100} \\
m_n(x_{n-1}, y_n) &= S_n(x_{n-1})\Big((B_n B_n^T)^{-1}\psi_{\theta}(x_{n-1},0) \\
&\quad + C_n^T (D_n D_n^T)^{-1} \big(y_n - \phi_{\theta}(\psi_{\theta}(x_{n-1},0),0) + C_n \psi_{\theta}(x_{n-1},0)\big)\Big) \\
m_n(x_{n-1}, y_n) &= S_n(x_{n-1}) \left(10\psi_{\theta}(x_{n-1},0)+ \frac{\psi_{\theta}(x_{n-1},0)}{10}\left(y_n + \frac{\psi_{\theta}(x_{n-1},0)^2}{20}\right)\right)
\end{aligned}
$$
For $q_{\theta}(x_0|y_0)$ the linearisation just leads to the bootstrap proposal, as linearising the likelihood of $y_0$ around the prior mean leads to $q_{\theta}(x_0|y_0) \sim N(0,1)$ (i.e. the bootstrap in this case).

Using multinomial resampling and the above proposals then led to the following particle approximations for $p(x_n|y_{0:n})$ in @fig-1 and @fig-2, where the densities are estimated via the default 'density' function in R, which uses a Gaussian kernel and rule-of-thumb bandwidth selector based on the standard deviation of the data. The red dotted line represents the true value of the hidden state, demonstrating how the SIR particle filter performed relatively accurately, even with N=50 (though increasing samples increased accuracy).



```{r}
#| echo: False
#A single forward run of a SIR particle filter

#creates a matrix with N rows and n columns 
#each column will have the Xi sampled at time i 
#weights will all be assumed to be 1/N, as the samples will be in-
#-the matrix only after resampling

psi0 <- function(x, n){
  return (0.5*x + (25*x)/(1+x^2) + 8*cos(1.2*n))
}

proposal <- function(x, y, n){
  psi <- psi0(x, n)
  Sn <- (10 + (psi^2)/100)^(-1)
  mn <- Sn * (10*psi + (psi/10) * (y + (psi^2)/20 ))
  return(list(mn = mn, Sn = Sn))
}

SIR <- function(N){
  samples <- matrix(rep(0,201*N), nrow=N, ncol=201)
  ancestors <- matrix(0, nrow=N, ncol=201)
  
  ## Sampling first samples
  
  samples0 <- rep(0,N)
  weights0 <- rep(0,N)
  
  for (i in 1:N){
    samples0[i] <- rnorm(1, mean=0, sd=1)
    #compute unnormalised weights- target over proposal
    #in bootstrap case this is just g
    weights0[i] <- dnorm(y[1], mean=(samples0[i]^2)/20, sd=1)
  }
  
  # normalise weights
  weights0 <- weights0/sum(weights0)
  
  #Resample (Multinomial Resampling)

  resamples <- sample(1:N, size = N, replace = TRUE, prob = weights0)
  ancestors[,1] <- resamples
  resamples0 <- rep(0,N)
  
  
  for (i in 1:N){
    resamples0[i] <- samples0[resamples[i]]
  }
  
  #update matrix
  samples[,1] <- resamples0
  
  ## Sampling remaining 199*N samples
  for (k in 2:201){
    samplesk <- rep(0,N)
    weightsk <- rep(0,N)
    
    
    for (i in 1:N){
      Xkprev <- samples[,k-1][i]
      prop <- proposal(x=Xkprev, y=y[k], n=k-1)
      samplesk[i] <- rnorm(1, mean=prop$mn, sd=sqrt(prop$Sn))
      
      fx <- psi0(Xkprev, k-1)
      transition <- dnorm(samplesk[i], mean=fx, sd=sqrt(0.1))
      likelihood <- dnorm(y[k], mean=(samplesk[i]^2)/20, sd=1)
      proposal_density <- dnorm(samplesk[i], mean=prop$mn, sd=sqrt(prop$Sn))
      
      
      
      
      log_w <- log(transition) + log(likelihood) - log(proposal_density)
      weightsk[i] <- exp(log_w)
    }
    #normalise weights
    weightsk <- weightsk/sum(weightsk)
    
    #Multinomial Resampling
    resamples <- sample(1:N, size=N, replace=TRUE, prob=weightsk)
    ancestors[,k] <- resamples
    resamplesk <- rep(0,N)
    
    for (i in 1:N){
      resamplesk[i] <- samplesk[resamples[i]]
    }
    
    #update matrix
    samples[,k] <- resamplesk
  }
  return(list(samples = samples, ancestors = ancestors))
}

SIR50 <- SIR(50)
sir50 <- SIR50$samples

SIR500 <- SIR(500)
sir500 <- SIR500$samples
```

```{r}
#| echo: False
#| label: fig-1
#| fig-cap: "N=50 SIR particle filter approximations"
#| fig-width: 12
#| fig-height: 4

times = c(2,51,101,151,200)
par(mfrow = c(1, length(times)))

for (n in times) {
  d <- density(sir50[, n])
  plot(d, lwd = 2, col = "blue",
       main = paste("n =", n-1),
       xlab = "x", ylab = "Density")
  abline(v = x[n], col = "red", lwd = 2, lty = 2)
}

```

```{r}
#| echo: False
#| label: fig-2
#| fig-cap: "N=500 SIR particle filter approximations"
#| fig-width: 12
#| fig-height: 4

times = c(2,51,101,151,200)
par(mfrow = c(1, length(times)))

for (n in times) {
  d <- density(sir500[, n])
  plot(d, lwd = 2, col = "blue",
       main = paste("n =", n-1),
       xlab = "x", ylab = "Density")
  abline(v = x[n], col = "red", lwd = 2, lty = 2)
}

```




### 2. Path Degeneracy

Plotting the final simulated trajectories $\bar{X}^i_{0:200}$ of the SIR algorithm in @fig-3 illustrates the path degeneracy phenomenon well.


```{r}
#| echo: False
#| label: fig-3
#| fig-cap: "Path Degeneracy Plot"


anc <- SIR50$ancestors
paths <- matrix(rep(0,201*50), nrow=50, ncol=201)
paths[,201] <- sir50[,201]


idx <- 201
while (idx>1){
  for (i in 1:50){
    ancestor <- sir50[anc[i,idx], idx-1]
    paths[i,idx-1] <- ancestor
      
  }
  idx <- idx-1
}


matplot(t(paths), type = "l", lty = 1, col = rgb(0, 0, 0, 0.01),
        xlab = "Time", ylab = "Value")




```

# Code Appendix {#sec-code-appendix}

```{r ref.label=knitr::all_labels()}
#| echo: true
#| eval: false
#| code-fold: true

```











