---
title: "Advanced Simulation Summative Coursework"
subtitle: 'MSc in Statistics 2025/26, Imperial College London'
author: "06051710"
format:
  html:
    toc: true
    highlight: tango
    self-contained: true
    df-print: paged
  pdf: default
format-links: false
editor: 
  markdown: 
    wrap: 72
---

```{=html}
<style type="text/css">
h1{
  font-size: 24pt;
}
h2{
  font-size: 18pt;
}
body{
  font-size: 12pt;
}
</style>
```

```{r setup, include = FALSE, tidy=TRUE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
include_solutions <- TRUE
```

```{r setup2, include=FALSE, tidy=TRUE}
set.seed(8)
require(rmarkdown)
require(knitr)
require(kableExtra)
# Put any library imports and other preamble here.
```

# Question 1: Filtering and smoothing for a nonlinear state space model

```{r}
#| echo: False

######
#Question 1a
######

#Simulating the dataset
x <- rep(0,201)
y <- rep(0,201)
x[1] <- rnorm(1)
y[1] <- (x[1]^2)/20 + rnorm(1)

for (i in 2:201){
  x[i] <- 0.5*x[i-1] + (25*x[i-1])/(1+(x[i-1])^2) + 8*cos(1.2*(i-1))+rnorm(1, mean=0, sd=sqrt(0.1))
  y[i] <- (x[i]^2)/20 + rnorm(1)
}
```

## 1a. Sampling from the SIR particle filter

For this problem, I chose as my importance density $q(x_n|y_n, x_{n-1})$ an approximation of the optimal proposal (that is $p(x_n|y_n, x_{n-1})$), by linearisations similar to the extended Kalman Filter (EKF). Although a bootstrap proposal would be easier, I felt the density $g(y_n|x_n)$ was relatively informative (or 'peaky'), which meant a more informative proposal seemed appropriate. Additionally, the notes suggest the EKF can give reasonable answers for models where the non-linearity is independent of the noise, which is the case for our model, as neither $V_n$ or $W_n$ appear in non-linear functions. 

$X_n$ can be linearly approximated as  $X_n = a_n + A_nX_{n-1} + B_nZ_n$ (where $Z_n \sim N(0,1)$) and $Y_n$ can be approximated as $Y_n = c_n +C_nX_n + W_n$ (where $W_n \sim N(0,1)$), both by a classic Taylor expansion. This leads to $B_n = \sqrt{0.1}$, $C_n = \frac{\psi_{\theta}(X_{n-1},0)}{10}$ and $D_n=1$  (all other variables are not relevant for the proposal). The approximation of the optimal proposal^[this result is taken from the notes] then takes the form $q(x_n|y_n, x_{n-1}) \propto N(m_n(x_{n-1}, y_n), S_n(x_{n-1}))$, where $m_n$ and $S_n$ are defined in @eq-1.

$$
\begin{aligned}
S_n^{-1}(x_{n-1}) &= (B_n B_n^T)^{-1} + C_n^T (D_n D_n^T)^{-1} C_n \\
S_n^{-1}(x_{n-1}) &= 10 + \frac{\psi_{\theta}(x_{n-1},0)^2}{100} \\
m_n(x_{n-1}, y_n) &= S_n(x_{n-1})\Big((B_n B_n^T)^{-1}\psi_{\theta}(x_{n-1},0) \\
&\quad + C_n^T (D_n D_n^T)^{-1} \big(y_n - \phi_{\theta}(\psi_{\theta}(x_{n-1},0),0) + C_n \psi_{\theta}(x_{n-1},0)\big)\Big) \\
m_n(x_{n-1}, y_n) &= S_n(x_{n-1}) \left(10\psi_{\theta}(x_{n-1},0)+ \frac{\psi_{\theta}(x_{n-1},0)}{10}\left(y_n + \frac{\psi_{\theta}(x_{n-1},0)^2}{20}\right)\right)
\end{aligned}
$$ {#eq-1}

For $q_{\theta}(x_0|y_0)$ the linearisation just leads to the bootstrap proposal, as linearising the likelihood of $y_0$ around the prior mean leads to $q_{\theta}(x_0|y_0) \sim N(0,1)$ (i.e. the bootstrap in this case).

Using multinomial resampling and the above proposals then led to the following particle approximations for $p(x_n|y_{0:n})$ in @fig-1 and @fig-2, where the densities are estimated via the default 'density' function in R, which uses a Gaussian kernel and rule-of-thumb bandwidth selector based on the standard deviation of the data. The red dotted line represents the true value of the hidden state, demonstrating how the SIR particle filter performed relatively accurately, even with N=50 (though increasing samples increased accuracy).



```{r}
#| echo: False
#A single forward run of a SIR particle filter

#creates a matrix with N rows and n columns 
#each column will have the Xi sampled at time i 
#weights will all be assumed to be 1/N, as the samples will be in-
#-the matrix only after resampling

psi0 <- function(x, n){
  return (0.5*x + (25*x)/(1+x^2) + 8*cos(1.2*n))
}

proposal <- function(x, y, n){
  psi <- psi0(x, n)
  Sn <- (10 + (psi^2)/100)^(-1)
  mn <- Sn * (10*psi + (psi/10) * (y + (psi^2)/20 ))
  return(list(mn = mn, Sn = Sn))
}

SIR <- function(N){
  samples <- matrix(rep(0,201*N), nrow=N, ncol=201)
  unweighted <- matrix(rep(0,201*N), nrow=N, ncol=201)
  sample_weights <- matrix(rep(0,201*N), nrow=N, ncol=201)
  ancestors <- matrix(0, nrow=N, ncol=201)
  
  ## Sampling first samples
  
  samples0 <- rep(0,N)
  weights0 <- rep(0,N)
  
  for (i in 1:N){
    samples0[i] <- rnorm(1, mean=0, sd=1)
    #compute unnormalised weights- target over proposal
    #in bootstrap case this is just g
    weights0[i] <- dnorm(y[1], mean=(samples0[i]^2)/20, sd=1)
  }
  
  # normalise weights
  weights0 <- weights0/sum(weights0)
  unweighted[,1] <- samples0
  sample_weights[,1]<-weights0
  
  #Resample (Multinomial Resampling)

  resamples <- sample(1:N, size = N, replace = TRUE, prob = weights0)
  ancestors[,1] <- resamples
  resamples0 <- rep(0,N)
  
  
  for (i in 1:N){
    resamples0[i] <- samples0[resamples[i]]
  }
  
  #update matrix
  samples[,1] <- resamples0
  
  ## Sampling remaining 199*N samples
  for (k in 2:201){
    samplesk <- rep(0,N)
    weightsk <- rep(0,N)
    
    
    for (i in 1:N){
      Xkprev <- samples[,k-1][i]
      prop <- proposal(x=Xkprev, y=y[k], n=k-1)
      samplesk[i] <- rnorm(1, mean=prop$mn, sd=sqrt(prop$Sn))
      
      fx <- psi0(Xkprev, k-1)
      transition <- dnorm(samplesk[i], mean=fx, sd=sqrt(0.1))
      likelihood <- dnorm(y[k], mean=(samplesk[i]^2)/20, sd=1)
      proposal_density <- dnorm(samplesk[i], mean=prop$mn, sd=sqrt(prop$Sn))
      
      
      
      
      log_w <- log(transition) + log(likelihood) - log(proposal_density)
      weightsk[i] <- exp(log_w)
    }
    #normalise weights
    weightsk <- weightsk/sum(weightsk)
    unweighted[,k] <- samplesk
    sample_weights[,k] <- weightsk
    
    #Multinomial Resampling
    resamples <- sample(1:N, size=N, replace=TRUE, prob=weightsk)
    ancestors[,k] <- resamples
    resamplesk <- rep(0,N)
    
    for (i in 1:N){
      resamplesk[i] <- samplesk[resamples[i]]
    }
    
    #update matrix
    samples[,k] <- resamplesk
  }
  return(list(samples = samples, ancestors = ancestors, unweighted=unweighted, sample_weights=sample_weights))
}

SIR50 <- SIR(50)
sir50 <- SIR50$samples
unweighted_sir50 <- SIR50$unweighted
weights_sir50 <- SIR50$sample_weights

system.time({
  SIR500 <- SIR(500)
})
sir500 <- SIR500$samples
unweighted_sir500 <- SIR500$unweighted
weights_sir500 <- SIR500$sample_weights
```

```{r}
#| echo: False
#| label: fig-1
#| fig-cap: "N=50 SIR particle filter approximations"
#| fig-width: 12
#| fig-height: 4

times = c(2,51,101,151,200)
par(mfrow = c(1, length(times)))

for (n in times) {
  d <- density(sir50[, n])
  plot(d, lwd = 2, col = "blue",
       main = paste("n =", n-1),
       xlab = "x", ylab = "Density")
  abline(v = x[n], col = "red", lwd = 2, lty = 2)
}

```

```{r}
#| echo: False
#| label: fig-2
#| fig-cap: "N=500 SIR particle filter approximations"
#| fig-width: 12
#| fig-height: 4

times = c(2,51,101,151,200)
par(mfrow = c(1, length(times)))

for (n in times) {
  d <- density(sir500[, n])
  plot(d, lwd = 2, col = "blue",
       main = paste("n =", n-1),
       xlab = "x", ylab = "Density")
  abline(v = x[n], col = "red", lwd = 2, lty = 2)
}

```

### 1b. Sampling using a particle smoothing algorithm

For this question I chose to use a Forward Filtering Backwards Smoothing (FFBSm) algorithm, as I felt it was the best option I know of for particle smoothing problems (and did not require any hyperparameter tuning). It led to the filter approximations in @fig-8 and @fig-9 (via the same smooth kernel density estimate as in 1a) which again seemed accurate in predicting hidden states. The terminology 'smoother' for $p(x_n|y_{0:T})$ seems justified, as visually speaking the density estimates appear 'smoother' than the SIR equivalents, likely due to using more information for their predictions.

```{r}
#| echo: False
#| output: False

######
#Question 1b
######

ffbsm <- function(samples, weights){
  N <- nrow(samples)
  T <- ncol(samples)
  
  back_weights <- matrix(0, nrow=N, ncol=T)
  back_weights[,T] <- weights[,T]
  
  for (k in (T-1):1){
    
    x_prev <- samples[,k]
    x_next <- samples[,k+1]
    
    # Transition matrix: [i, j] = f(x_next[j] | x_prev[i])
    trans_mat <- outer(
      x_prev, x_next,
      Vectorize(function(xp, xn) {
        dnorm(xn, mean = psi0(xp, k), sd = sqrt(0.1))
      })
    )
    
    # denom_j = sum_l weights[l,k] * f(x_next[j] | x_prev[l])
    denom <- as.vector(weights[,k] %*% trans_mat)
    
    # Avoid division by zero
    denom[denom == 0] <- 1e-12
    
    # Compute sum over j efficiently
    # back_weights[j,k+1] / denom[j]
    ratio <- back_weights[,k+1] / denom
    
    # sum_j ratio[j] * f(x_next[j] | x_prev[i])
    back_weights[,k] <- weights[,k] * (trans_mat %*% ratio)
    
    # normalise
    back_weights[,k] <- back_weights[,k] / sum(back_weights[,k])
  }
  
  return(back_weights)
}

system.time({
  ffbsm_weights1 <- ffbsm(unweighted_sir50, weights_sir50)
})

system.time({
  ffbsm_weights2 <- ffbsm(unweighted_sir500, weights_sir500)
})

```

```{r}
#| echo: False
#| label: fig-8
#| fig-cap: "N=50 FFBSm Approximation"
#| fig-width: 12
#| fig-height: 4
#| warning: False



times = c(2,51,101,151,200)
par(mfrow = c(1, length(times)))

for (n in times) {
  d <- density(unweighted_sir50[,n],
               weights = ffbsm_weights1[,n])
  plot(d, lwd = 2, col = "blue",
       main = paste("n =", n-1),
       xlab = "x", ylab = "Density")
  abline(v = x[n], col = "red", lwd = 2, lty = 2)
}
```

```{r}
#| echo: False
#| label: fig-9
#| fig-cap: "N=500 FFBSm Approximation"
#| fig-width: 12
#| fig-height: 4
#| warning: False



times = c(2,51,101,151,200)
par(mfrow = c(1, length(times)))

for (n in times) {
  d <- density(unweighted_sir500[,n],
               weights = ffbsm_weights2[,n])
  plot(d, lwd = 2, col = "blue",
       main = paste("n =", n-1),
       xlab = "x", ylab = "Density")
  abline(v = x[n], col = "red", lwd = 2, lty = 2)
}
```



### 2a. Path Degeneracy for the SIR filter

The SIR filter has clear path degeneracy (for N=50). This is not clear if the paths of all particles are plotted due to the scale of the data, however, it is clearly demonstrated in @fig-4 where paths from n=190 onwards are plotted and compared to the paths from n=0 to 20. In fact, there is only a single path when going back from n=154^[code for this is in the appendix], suggesting path degeneracy is a strong issue for the SIR filter.


```{r}
#| echo: False

######
#Question 2a
######

path_generator <- function(anc, samples){
  paths <- matrix(0, nrow=length(samples[,1]), ncol=201)
  paths[,201] <- samples[,201]
  indices <- 1:length(samples[,1])
  idx <- 201
  while (idx > 1) {
    indices <- anc[indices, idx]
    paths[, idx-1] <- samples[indices, idx-1]
    idx <- idx - 1
  }
  return(paths)
}
```

```{r}
#| echo: False
#| label: fig-4
#| fig-cap: "Visualising the start and end of the paths of the SIR filter"
#| fig-width: 12
#| fig-height: 4


paths50 <- path_generator(SIR50$ancestors, SIR50$samples)
par(mfrow=c(1,2))

matplot(t(paths50[,0:20]), type = "l", lty = 1, col = rgb(0, 0, 0, 0.1),
        xlab = "Time", ylab = "Value")
matplot(190:200, t(paths50[,190:200]), type = "l", lty = 1,
        col = rgb(0, 0, 0, 0.1),
        xlab = "Time", ylab = "Value")
```
```{r}
#| echo: False
#| output: False

length(unique(paths50[,201]))
length(unique(paths50[,200]))
length(unique(paths50[,170]))
length(unique(paths50[,160]))
length(unique(paths50[,154]))
length(unique(paths50[,100]))
length(unique(paths50[,20]))
length(unique(paths50[,1]))

```

### 2b. Path Degeneracy for the FFBSm Algorithm

Although it is not possible to view 'paths' generated by the FFBSm Algorithm, as ultimately the forward run is identical to the generic SIR algorithm, we can still see the improvement in addressing path degeneracy when calculating particle approximations of $p(x_n|y_{0:T})$. For example, as explained previously, only one unique particle exists among all paths when calculating $p(x_{n}|y_{0:T})$ for $n \leq 154$ with the SIR filter, though in @fig-2b we can see there are almost 20 unique particles even for $n \leq 10$ (though this does vary depending on n).

```{r}
#| echo: False
#| label: fig-2b
#| fig-cap: "Visualising the number of unique particles in the FFBSm approximation"
#| fig-width: 12
#| fig-height: 4

######
#Question 2b
######

unique_particles <- function(samples, weights){
  # samples: 50 x 201 matrix
  # weights: same size
  
  n_particles <- nrow(samples)
  n_steps <- ncol(samples)
  unique_counts <- numeric(n_steps)
  
  for (k in 1:n_steps){
    weightsk <- weights[, k]
    resamples <- sample(1:n_particles, size = n_particles,
                        replace = TRUE, prob = weightsk)
    resampled_values <- samples[resamples, k]
    unique_counts[k] <- length(unique(resampled_values))
  }
  return(unique_counts)
}

plot(1:201, unique_particles(sir50, ffbsm_weights1),
     xlab="Time",
     ylab="Number of Unique particles")

```


### 3. Particle Approximations for $E[X_n|y_{0:T}]$ and $E[X_n|y_{0:n}]$

To work out the SIR approximation of $E[X_n|y_{0:T}]$ I used the paths that I previously calculated to illustrate path degeneracy and calculated their mean. @fig-31 and @fig-32 display the true value of the hidden state at each time n in red, alongside both particle approximations of $E[X_n|y_{0:T}]$ with the FFBSm approximation in green and the SIR approximation in black, as well as the filter approximation of $E[X_n|y_{0:n}]$ in blue. It is clear most approximations are very similar in value calculated, though the variance of the filter is much greater than both approaches used to calculate $p(x_n|y_{0:T})$. 

```{r}
#| echo: False
#| label: fig-31
#| fig-cap: "Filter vs Smoothing Estimates and Variance N=50"
#| fig-width: 12
#| fig-height: 4

######
#Question 3 Plot 1
######

smooth_mean <- function(samples, back_weights){
  colSums(samples * back_weights)
}


EX_smooth <- colMeans(paths50)
EX_smooth_ffbsm <- smooth_mean(unweighted_sir50, ffbsm_weights1)
EX_filter <- colMeans(SIR50$samples)

par(mfrow = c(1, 2))

plot(1:201, EX_filter, type="l", lwd=2,
     xlab="Time", ylab="Expectation",
     ylim = range(c(EX_filter, EX_smooth, x)),
     col="blue")

lines(1:201, EX_smooth_ffbsm, col="green", lwd=2)
lines(1:201, EX_smooth, col="black", lwd=2)
lines(1:201, x, col="red", lty=2, lwd=2)

var_calc <- function(samples){
  apply(samples, 2, var)
}

VAR_filter_50 <- var_calc(sir50)
VAR_smooth_paths50 <- var_calc(paths50)

smooth_var <- function(samples, weights){
  means <- colSums(samples * weights)
  vars <- colSums(weights * (samples - matrix(means, 
                    nrow=nrow(samples), 
                    ncol=ncol(samples), 
                    byrow=TRUE))^2)
  return(vars)
}

VAR_smooth_ffbsm_50 <- smooth_var(unweighted_sir50, ffbsm_weights1)

plot(1:201, VAR_filter_50, type="l", lwd=2,
     xlab="Time", ylab="Variance",
     ylim = range(c(VAR_filter_50,
                    VAR_smooth_paths50,
                    VAR_smooth_ffbsm_50)),
     col="blue")

lines(1:201, VAR_smooth_ffbsm_50, col="green", lwd=2)
lines(1:201, VAR_smooth_paths50, col="black", lwd=2)


legend("topright",
       legend=c("Filter Var", "SIR Smooth Var", "FFBSm Smooth Var"),
       col=c("blue","black","green"),
       lwd=2)

```



```{r}
#| echo: False
#| label: fig-32
#| fig-cap: "Filter vs Smoothing Estimates and Variance N=500"
#| fig-width: 12
#| fig-height: 4

######
#Question 3 plot 2 (N=500)
######

paths500 <- path_generator(SIR500$ancestors, SIR500$samples)


EX_smooth500 <- colMeans(paths500)
EX_smooth_ffbsm500 <- smooth_mean(unweighted_sir500, ffbsm_weights2)
EX_filter500 <- colMeans(SIR500$samples)

par(mfrow = c(1, 2))
plot(1:201, EX_smooth500, type="l", lwd=2,
     xlab="Time", ylab="Expectation",
     ylim = range(c(EX_filter500, EX_smooth500, x)),
     col="blue")

lines(1:201, EX_smooth_ffbsm500, col="green", lwd=2)
lines(1:201, EX_smooth500, col="black", lwd=2)
lines(1:201, x, col="red", lty=2, lwd=2)

VAR_filter_500 <- var_calc(sir500)
VAR_smooth_paths500 <- var_calc(paths500)
VAR_smooth_ffbsm_500 <- smooth_var(unweighted_sir500, ffbsm_weights2)

plot(1:201, VAR_filter_500, type="l", lwd=2,
     xlab="Time", ylab="Variance",
     ylim = range(c(VAR_filter_500,
                    VAR_smooth_paths500,
                    VAR_smooth_ffbsm_500)),
     col="blue")

lines(1:201, VAR_smooth_ffbsm_500, col="green", lwd=2)
lines(1:201, VAR_smooth_paths500, col="black", lwd=2)


legend("topright",
       legend=c("Filter Var", "SIR Smooth Var", "FFBSm Smooth Var"),
       col=c("blue","black","green"),
       lwd=2)


```

### 4. Comments on results

Visually speaking, when comparing the SIR approximation of $p(x_n|y_{0:n})$ to the FFBSm approximation of $p(x_n|y_{0:T})$ the FFBSm approach yields a much 'smoother' density, though I would not say the results seem more accurate than the SIR approximation in predicting the hidden states. The 'smoothness' is also reflected in the increased variance of the filter compared to the smoother, seen in question 3, though again, it seems all algorithms are still able to calculate $E[X_n|y_{0:T}]$ to similar degrees of accuracy.

Smoothing uses more information than filtering and is offline, whereas filtering is online. Though online algorithms are more useful when real-time parameter updates are sought after, having well-designed smoothing algorithms is necessary for parameter estimation.

Increasing N to 500 from 50 uniformly improved estimation accuracy in calculating the hidden state for all algorithms. I took T=200, though simulated 201 samples from the true model, as there was also an $x_0$ and $y_0$ term. The forward run of the SIR filter used multinomial resampling, as the basic version seemed to perform decently in terms of accuracy, though more advanced resampling techniques *could* have reduced the variance of estimates (they likely still would have been longer than the smoothers).

The most computationally expensive algorithm by far was the FFBSm implementation with N=500, as it has a cost of $O(N^2T)$, leading to the largest runtime. On my device, the implementation of the forward run of the SIR algorithm (with N=500) took 0.65 seconds compared to 104 seconds for the implementation of the FFBSm algorithm. 




# Code Appendix {#sec-code-appendix}

```{r ref.label=knitr::all_labels()}
#| echo: true
#| eval: false
#| code-fold: true

```











